---
title: "Validating Clusters in Tuberculosis Monitoring Indicators - using Thresher"
author: 'null'
date: '2018-04-12'
description: "Using fingertipsR, principal component analysis, and partitioning around medoids to identify clusters of counties based on Tuberculosis monitoring indicators"
slug: thresh-cluster-england-tb
draft: yes
tags: ["data analysis", "data visualisation", "rstats", "TB", "PHE", "infectious disease", "clustering", "thresher"]
categories: ["R"]
twitterImg: "img/fingertips/storyboard-fingertips-tb-clust.png"
---

```{r knitr-opts, echo = FALSE}
knitr::opts_chunk$set(cache = TRUE, dpi = 330, 
                      fig.height = 8, fig.width = 8,
                      warning = FALSE)
```

## Introduction

- Previous post using `fingertipsR` to extract Tuberculosis monitoring indicators for England
- Used PCA and PAM clustering to determine the optimal number of clusters in the data and then compared these clusters
- This post using new package Thresher, explore package functionality, compare results to previous findings
- What is the impact of excluding Incidence rates from the analysis?

## Packages

The first step is to load the packages required for the analysis, we do this using the fantastic `pacman` package.

```{r get-packages, message = FALSE}
if (!require(pacman)) install.packages("pacman"); library(pacman)
p_load("tidyverse")
p_load("purrr")
p_load_gh("thomasp85/patchwork", dependencies = TRUE)
```

## Data

```{r get-prev-post-data}
tb_monitoring_indicators <- readRDS("../../static/data/fingertips/tb-pam-clusters.rds")

tb_monitoring_indicators
```

## Analysis
