---
title: getTBinR 0.5.7 now on CRAN - Tuberculosis reports and summary plots
author: 'null'
date: '2019-01-22'
slug: gettbinr-5-7
twitterImg: img/getTBinR/release-5-7.png
description: "getTBinR version 0.5.7 now with a new plotting function to quickly show regional and global metrics + a new country level Tuberculosis report generating function. Tweet an instance of you using the package at me and potentially recieve a hexsticker!"
categories: ["R"]
tags: ["data analysis", "data visualisation", "rstats", "TB", "WHO", "getTBinR", "infectious disease"]
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, dpi = 320, 
                      fig.height = 8, fig.width = 8,
                      eval = TRUE)
```

```{r get-packages, message = FALSE, include = FALSE, echo = FALSE}
if (!require(pacman)) install.packages("pacman"); library(pacman)
p_load_gh("seabbs/getTBinR")
## Load additional packages
p_load("dplyr") # For data munging 
p_load("tidyr")
p_load("rlang")
p_load("ggplot2")
```

[`getTBinR 0.5.7`](https://www.samabbott.co.uk/getTBinR/) is now on CRAN and should be available on a mirror near you shortly! This update mainly focussed on building out new country level Tuberculosis (TB) report functionality but along the way this led to a new summary plotting function that quickly and easily shows TB trends across regions and globally. I also had some fun developing a hexsticker (Tweet at me with something you made using the package to get a physical version - whilst my postage money lasts...), reducing the dependencies with [`itdepends`](https://github.com/jimhester/itdepends) and [`pkgnet`](https://uptakeopensource.github.io/pkgnet/index.html) and dealing with some breaking changes from an uncoming `dplyr` update (my own fault for missing a function import).

The full changelog is below along with an example of the country level TB report generated for the UK (generate a report on the country of your choice using `getTBinR::render_country_report(country = "United Kingdom", save_dir = ".")`).

## Feature updates

* Added support for annual_change to summarise_tb_burden and added validating tests.
* Added support for rates and proportions to summarise_tb_burden and added validating tests.
* Added a new function - plot_tb_burden_summary. Function wraps summarise_tb_burden and allows all in one summary plotting. Inspired by this case study.
* Added a rmarkdown parameterised country level report on TB.
* Added a report generating button to the dashboard generated by run_tb_dashboard.
* Added render_country_report to generate a TB report for a given country.
* Tweaked map_tb_burden to not use geom_path for country outlines.
* Added a smooth argument to plot_tb_burden to allow smooth trend lines to be plotted (derived using ggplot2::geom_smooth).
* Tweaked line thickness in plot_tb_burden to improve plot appearance.
* Added legend argument to all plotting functions to allow control of the legend appearance.

## Package updates

* Added script to generate hexsticker
* Added hexsticker to README
* Added DOI link to Zenodo.
* Updated tests to account for dplyr 8.0 release and vdiffr updates.
* Added itdepends to package report functionality.

## Example: United Kingdom Tuberculosis Report


```{r report-setup, results = "hide"}
## Load the package
library(getTBinR)
## Load additional packages
library(dplyr) # For data munging 
library(tidyr)
library(rlang)
library(ggplot2)

## Get the data
tb <- get_tb_burden(verbose = FALSE)
## Get the data dictionary
dict <- get_data_dict(verbose = FALSE)
##Assign parameters - these are set in the YAML within the package
country <- "United Kingdom"
interactive <- FALSE
```

## TB incidence rates

```{r get country-stats, results = "hide"}
metric_summary <- function(df = NULL, target_country = NULL, metric = NULL) {
  
  target_country <- df$country[grepl(target_country, df$country)] %>% 
    unique %>% 
    first
    
  ##  Set up metric with confidence intervals
  metric <- enquo(metric)
  metric_lo <- sym(paste0(quo_name(metric), "_lo"))
  metric_hi <- sym(paste0(quo_name(metric), "_hi"))
  
  ## Filter for the country of interest
  country_df <- df %>% 
  filter(country %in% target_country)
  
## Most up to date year of incidence data
recent_inc <- country_df %>% 
  drop_na(!!metric) %>% 
  filter(year == max(year)) %>% 
  select(!!metric, !!metric_lo, !!metric_hi, year, g_whoregion) %>% 
  mutate(inc_rate = paste0(!!metric, " (", !!metric_lo, " - ", !!metric_hi, ")"))
## Country rank
ranked_countries_inc <- df %>% 
  filter(year == recent_inc$year) %>% 
  arrange(desc(!!metric)) %>% 
  mutate(rank = 1:n())
## World rank
target_rank_world <- ranked_countries_inc %>% 
  filter(country == target_country) %>% 
  pull(rank)
## Region rank
target_rank_region <- ranked_countries_inc %>% 
  filter(g_whoregion %in% recent_inc$g_whoregion) %>% 
  mutate(rank = 1:n()) %>% 
  filter(country == target_country) %>%
  pull(rank)
## Summarise annual change
country_change <- summarise_tb_burden(metric = quo_name(metric),
                                      stat = "mean",
                                      countries = target_country,
                                      compare_to_region = FALSE,
                                      compare_to_world = FALSE,
                                      compare_all_regions = FALSE,
                                      annual_change = TRUE,
                                      verbose = FALSE) %>% 
  filter(year > (max(year) - 10)) %>% 
  summarise(change = mean(!!metric, na.rm = FALSE)) %>%
  mutate(change = round(change * 100, 1) %>% 
           paste0(., "%")) %>% 
  pull(change)
out <- list(recent_inc$year[1], recent_inc$inc_rate[1],
            target_rank_world, target_rank_region,
            country_change)
names(out) <- c("year", "metric", "world_rank", "region_rank", "avg_change")
out <- ifelse(is.na(out), "(Missing)", out)
return(out)
}
  
  
inc_sum <- metric_summary(tb, country, e_inc_100k)
```

In `r inc_sum$year` `r country` had an estimated Tuberculosis incidence rate of `r inc_sum$metric` per 100,000 people making it number `r inc_sum$world_rank` in the world and number `r inc_sum$region_rank` regionally. In the last 10 years this has changed by `r inc_sum$avg_change` on average each year. 

### Regional and Global Trends Comparision

```{r}
plot_tb_burden_summary(countries = country,
                       metric_label = "e_inc_100k",
                       compare_to_world = TRUE, 
                       compare_to_region = TRUE,
                       compare_all_regions = FALSE,
                       annual_change = FALSE,
                       facet = "Area",
                       scales = "free_y",
                       legend = "none",
                       interactive = interactive,
                       verbose = FALSE)
```

## Rates Regional Breakdown

```{r, fig.height = 8, fig.width = 8 }
plot_tb_burden_overview(countries = country,
                        compare_to_region = TRUE,
                        interactive = interactive,
                        verbose = FALSE)
```

## Case Detection Rates (CDR)

```{r, include = FALSE, results = "hide"}
cdr_sum <- metric_summary(tb, country, c_cdr)
```

`r country` had an estimated case detection rate of `r cdr_sum$metric`% in `r cdr_sum$year` making it number `r cdr_sum$world_rank` in the world (with number 1 having the highest CDR) and number `r cdr_sum$region_rank` regionally. In the last 10 years this has changed by `r cdr_sum$avg_change` on average each year. 

### Regional Breakdown

```{r, fig.height = 8, fig.width = 8}
plot_tb_burden_overview(metric = "c_cdr",
                        countries = country,
                        compare_to_region = TRUE,
                        interactrive = interactive,
                        verbose = FALSE)
```

## TB mortality rates - excluding HIV


```{r, include = FALSE, results = "hide"}
mort_exc_hiv_sum <- metric_summary(tb, country, e_mort_exc_tbhiv_100k)
```

In `r mort_exc_hiv_sum$year` `r country` had an estimated Tuberculosis mortality rate (excluding HIV) of `r mort_exc_hiv_sum$metric`  per 100,000 people making it number `r mort_exc_hiv_sum$world_rank` in the world and number `r mort_exc_hiv_sum$region_rank` regionally. In the last 10 years this has changed by `r mort_exc_hiv_sum$avg_change` on average each year. 

### Proportion of TB Cases that Died (excluding HIV) - Regional and Global Comparision

```{r, fig.height = 8, fig.width = 12}
plot_tb_burden_summary(metric = "e_mort_exc_tbhiv_num",
                       denom = "e_inc_num",
                       rate_scale = 100,
                       countries = country,
                       compare_to_region = TRUE,
                       compare_all_regions = FALSE,
                       interactive = interactive,
                       verbose = FALSE,
                       facet = "Area",
                       scales = "free_y",
                       legend = "none") +
  labs(y = "Proportion (%) of TB cases that died (excluding HIV)")
```

### Rates Regional Breakdown

```{r, fig.height = 8, fig.width = 8}
plot_tb_burden_overview(metric = "e_mort_exc_tbhiv_100k",
                        countries = country,
                        compare_to_region = TRUE,
                        interactrive = interactive,
                        verbose = FALSE)
```

## TB HIV related mortality rates

```{r, include = FALSE, results = "hide"}
mort_inc_hiv_sum <- metric_summary(tb, country, e_mort_tbhiv_100k)
```

In `r mort_inc_hiv_sum$year` `r country` had an estimated Tuberculosis mortality rate (related to HIV) of `r mort_inc_hiv_sum$metric`  per 100,000 people making it number `r mort_inc_hiv_sum$world_rank` in the world and number `r mort_inc_hiv_sum$region_rank` regionally. In the last 10 years this has changed by `r mort_inc_hiv_sum$avg_change` on average each year. 

### Proportion of TB Cases that Died (related to HIV) - Regional and Global Comparision

```{r, fig.height = 8, fig.width = 12}
plot_tb_burden_summary(metric = "e_mort_tbhiv_num",
                       denom = "e_inc_num",
                       rate_scale = 100,
                       countries = country,
                       compare_to_region = TRUE,
                       compare_all_regions = FALSE,
                       interactive = interactive,
                       verbose = FALSE,
                       facet = "Area",
                       scales = "free_y",
                       legend = "none") +
  labs(y = "Proportion (%) of TB cases that died (related to HIV)")
```

### Rates Regional Breakdown

```{r, fig.height = 8, fig.width = 8}
plot_tb_burden_overview(metric = "e_mort_tbhiv_100k",
                        countries = country,
                        compare_to_region = TRUE,
                        interactrive = interactive,
                        verbose = FALSE)
```

For other examples of using `getTBinR` to visualise the WHO TB data see [my](https://gist.github.com/seabbs) gists, previous blog [posts](https://www.samabbott.co.uk/tags/who/), and the [`getTBinR` website](https://www.samabbott.co.uk/getTBinR/).
